Path Analysis - Chapter 2

### 2015-12-21 sem_development.Rmd
Update rainfall data sources, rename some objects.
Saved data file 'data_201508.RData' holding the R data objects.

# Elevation and vegetation height are in feet, slope is degrees, and canopy density is the proportion of the cell covered by canopy (0-1)
plots_topo.shp <- readOGR("analysis/data/plot_gis/", "plots202_vegtopo")
plots_vegtopo <- plots_topo.shp@data
summary(plots_vegtopo)
plots_vegtopo <- select(plots_vegtopo, -X, -Y)
```
```{r load weather data, echo=FALSE}
# Temperature: rainy season hour summaries (long format for each year)
plots_temps <- read.csv("analysis/data/rainy_season_temps.csv")
summary(plots_temps)
plots_temps <- select(plots_temps, -X)

# Rainfall: interpolated rainfall for each season
plots_rain <- read.csv("analysis/data/plot202rain.csv")
summary(plots_rain)
plots_rain <- select(plots_rain, -cat)
# pairs(select(plots_rain, -PLOT_ID), 
#       lower.panel = panel.cor, 
#       diag.panel = panel.hist)
plots_rain <- rename(plots_rain, plotid = PLOT_ID)
```

@@

#### Plot contributing area vs. slope
This is a suggestion from Ryan to plot the contributing area vs. the slope for each of the plots in order to tease out how either influences infection and hosts. Decomposing TWI to determine which factor has a greater influence on disease dynamics. 
```{r decomposing TWI influence}
library(ggplot2)
summary(plots_vegtopo)
par(mfrow = c(3,1))
qplot(slope15m, flowacc15m, data = plots_env) +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m < 20), 
      main = "Flow Accumulation < 5") +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 20 & flowacc15m < 1000), main = "Flow Accumulation 5 - 1000") +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 1000), 
      main = "Flow Accumulation > 1000") +
      geom_text(aes(label = plotid))

@@

#### Examine the bay laurel data at individual and plot levels
```{r examine umca data}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.
summary(bay_laurel) # This has stem-level data on bay laurel.
```

I can use this to calculate plot level leaf count variables (sum, average) for each plot and year
```{r calculate plot level leaf counts}
plots_umca <- left_join(plots_umca, bay_laurel %>% filter(status == "Alive") %>% 
      group_by(plot, year) %>% 
      summarise(
            avg_lfct = mean(slc, na.rm = T),
            tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(plots_umca)
```

```{r examine lfct NAs}
filter(plots_umca, is.na(avg_lfct))
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm removing it manually from the `plots_umca` data frame
plots_umca <- plots_umca %>% filter(avg_lfct != "NaN")
summary(plots_umca)
```

@@

Now I also want to calculate some plot-level community variables from the vegetation data. First, understory richness from the vegetation transects.

@@

veg <- rbind(veg_us_2005, veg_us_2011)

unique(veg$year)
veg$year[veg$year < 2011] <- 2005 # Creates the two most complete veg sampling
veg <- veg %>% rename(us_species = UnderstorySp, plotid = PlotID)
unique(veg$us_species)

@@

woody_spp <- c(woody_spp, "ceanothus sp.", "cebe", "cecu", "frla", "garrya sp.", "gemo", "juca", "mafu", "mane", "mimulus sp", "pisa", "quag x quke", "qube", "quercus sp.", "quwi", "rhamnus sp.", "rhcr", "ribes sp.", "rica", "risa", "roca", "rosa sp.", "rubus sp.", "rudi", "salix sp.", "syal", "vaccinium sp.")

veg_sub <- filter(veg, veg$us_species %in% woody_spp)

@@

```{r calculate overstory richness}
summary(untagged_dbh)
untagged_dbh <- as.tbl(untagged_dbh %>% select(-X) %>% rename(plotid = PlotID, species = Species))
untagged_dbh
unique(untagged_dbh$year)
untagged_dbh$year[untagged_dbh$year < 2014] <- 2005 # Create two most complete groups
untagged_dbh$plotid <- tolower(untagged_dbh$plotid)

untagged_dbh %>% group_by(plotid, year) %>% summarise(untagged_rich = length(unique(species)))

stems <- as.tbl(stems)
summary(stems)
unique(stems$status)

stems <- stems %>% select(-X) %>% filter(status == "Alive" | status == "Dead")
unique(stems$species)
stems$species <- tolower(stems$species)
stems$species[stems$species == "quke x quag"] <- "quag x quke"
stems$plotid <- tolower(stems$plotid)

live_tagged_dbh <- stems %>% filter(status == "Alive") %>% group_by(plotid, year, species) %>% summarise(live_count = length(species))
unique(live_tagged_dbh$species)
os_richness <- rbind(live_tagged_dbh, select(untagged_dbh, plotid, species, year, live_count))

@@

Now I want to join the richness data set and the canopy cover estimates to the `plots_umca` data frame that currently has the environmental and leaf count/infection variables.

```{r join richness and umca data}
plots_umca
richness
plots_umca <- rename(plots_umca, plotid = plot)
plots_umca$plotid <- tolower(plots_umca$plotid)

plots_umca <- left_join(plots_umca, richness, by = c("plotid", "year"))
head(canopy_cover)
canopy_cover <- select(canopy_cover, plotid, year, pct_cvr)
plots_umca <- left_join(plots_umca, canopy_cover, by = c("plotid", "year"))
```

@@

plots_umca <- ungroup(plots_umca)

@@

```{r join umca and qusp data}
plots_umca
plot_qusp # right now this is broken down by oak species
summary(plot_qusp)
oak_sod <- plot_qusp %>% group_by(plot, year) %>%
      summarise(uninf_oak_ct = sum(uninfected_oak_ct), 
                inf_oak_ct = sum(infected_oak_ct), 
                tot_oak_ct = sum(c(uninfected_oak_ct, infected_oak_ct)))
oak_sod
oak_sod <- rename(oak_sod, plotid = plot)
oak_sod$plotid <- tolower(oak_sod$plotid)

oak_sod <- left_join(oak_sod, select(plots_umca, plotid, year, uninfected_bay_ct, infected_bay_ct, tot_bay, avg_lfct, tot_lfct), 
                     by = c("plotid", "year"))
summary(oak_sod) # the NA's for the bay laurel variables are legitimately zeroes b/c there are no bay in some of these plots
oak_sod$uninfected_bay_ct[is.na(oak_sod$uninfected_bay_ct)] <- 0
oak_sod$infected_bay_ct[is.na(oak_sod$infected_bay_ct)] <- 0
oak_sod$tot_bay[is.na(oak_sod$tot_bay)] <- 0

oak_sod <- left_join(oak_sod, plots_env, by = "plotid")
oak_sod <- left_join(oak_sod, plots_temps, 
                     by = c("plotid", "year" = "year"))
oak_sod

richness
oak_sod <- left_join(oak_sod, richness, by = c("plotid", "year"))
summary(oak_sod)
summary(filter(oak_sod, is.na(avg_tmax)))
filter(oak_sod, is.na(avg_tmax), year == 2005) # BUSH01 plot; I'm dropping that from the data set
oak_sod <- filter(oak_sod, plotid != "bush01")
filter(oak_sod, is.na(avg_tmax), year == 2004) # PONTI01 and SWEET01 were other early abandoned plots
oak_sod <- filter(oak_sod, plotid != "ponti01", plotid != "sweet01")
summary(filter(oak_sod, is.na(avg_tmax))) # Now the NAs in the temperature variables are just for year 2003 when we had severely incomplete temperature data
summary(oak_sod)
filter(oak_sod, is.na(easting))
oak_sod <- filter(oak_sod, plotid != "duer01")

@@

oak_sod <- left_join(oak_sod, canopy_cover, by = c("plotid", "year"))

@@

## Path Analysis Oak Infection 2005

Here I am using the `oak_sod` data set which will be the oak infection (binomial) as the terminal response variable.

@@

## Path Analysis Oak Infection - 2014
Now I am going to fit the same path model to data from 2011.

@@

```{r oak path 2}
oak_path2 <- DAG(
      tmi ~ elevation,
      os_rich ~ tmi,
      hrs_14_20 ~ elevation + d2c + avg_psi + tmi,
      tot_lfct ~ hrs_14_20 + os_rich + d2c + elevation,
      inf_oak_ct ~ tot_lfct + avg_psi + os_rich + hrs_14_20 + d2c + elevation
      )
isAcyclic(oak_path2)
bu_oak_path2 <- basiSet(oak_path2)
bu_oak_path2

@@

I also want to calculate some variables related to DBH, and in particular the "importance value" for each species. Importance value is calculated in the literature as:

`IV = 0.5 * (relative density / relative dominance)` where 
`Relative Density = # stems of species / total # of stems` and
`Relative Dominance = basal area of species / total basal area`

@@

stems_dbh <- rbind(tagged_dbh, untagged_dbh)

stems_dbh <- stems_dbh %>% group_by(plotid, year) %>% mutate(all_live_count = sum(live_count), all_live_dbh = sum(live_tot_dbh, na.rm = T))
summary(stems_dbh)
filter(stems_dbh, all_live_dbh > 800)

@@

>>>>

# Elevation and vegetation height are in feet, slope is degrees, and canopy density is the proportion of the cell covered by canopy (0-1)
plots_topo.shp <- readOGR("analysis/data/plot_gis/", "plots202_vegtopo")
plots_vegtopo <- plots_topo.shp@data
summary(plots_vegtopo)
plots_vegtopo <- select(plots_vegtopo, -cat) %>% rename(lon_utm = X, lat_utm = Y)
```
```{r load weather data, echo=FALSE}
# Temperature: rainy season hour summaries (long format for each year)
plots_temps <- read.csv("analysis/data/rainy_season_temps.csv")
summary(plots_temps)
plots_temps <- select(plots_temps, -X)

# Rainfall: interpolated/estimated rainfall for each season
load("~/Documents/soco_ppt/plot_ppt_estimates.RData")
summary(ppt_days)
summary(ppt_total)
# plots_rain <- read.csv("analysis/data/plot202rain.csv")
# summary(plots_rain)
# plots_rain <- select(plots_rain, -cat)
# pairs(select(plots_rain, -PLOT_ID), 
#       lower.panel = panel.cor, 
#       diag.panel = panel.hist)
plots_rain <- rename(plots_rain, plotid = PLOT_ID)
```

@@

#### Plot contributing area vs. slope
This is a suggestion from Ryan to plot the contributing area vs. the slope for each of the plots in order to tease out how either influences infection and hosts. Decomposing TWI to determine which factor has a greater influence on disease dynamics. 
```{r decomposing TWI influence}
library(ggplot2)
# summary(plots_vegtopo)
# par(mfrow = c(3,1))
qplot(slope15m, flowacc15m, data = plots_env) +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m < 20), 
      main = "Flow Accumulation < 20") +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 20 & flowacc15m < 1000), main = "Flow Accumulation 20 - 1000") +
      geom_text(aes(label = plotid))
qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 1000), 
      main = "Flow Accumulation > 1000") +
      geom_text(aes(label = plotid))

@@

#### Examine the bay laurel data at individual and plot levels
```{r examine umca data}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.
unique(plots_umca$plot)
umca_plots <- filter(plots_umca, plot != "BUSH01")
rm(plots_umca)
summary(bay_laurel) # This has stem-level data on bay laurel.
unique(bay_laurel$plot)
bay_laurel <- filter(bay_laurel, plot != "BUSH01")
```

I can use this to calculate plot level leaf count variables (sum, average) for each plot and year
```{r calculate plot level leaf counts}
umca_plots <- left_join(umca_plots, bay_laurel %>% filter(status == "Alive") %>% group_by(plot, year) %>%  summarise(
            avg_lfct = mean(slc, na.rm = T),
            tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(umca_plots)
```

```{r examine lfct NAs}
filter(umca_plots, is.na(avg_lfct))
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm removing it manually from the `plots_umca` data frame
umca_plots <- umca_plots %>% filter(avg_lfct != "NaN")
summary(umca_plots)
```

@@

#### Community vegetation data and variables
Now I also want to calculate some plot-level community variables from the vegetation data. First, understory richness from the vegetation transects.

@@

veg <- rbind(veg_us_2005, veg_us_2011)
unique(veg$year)
veg$year[veg$year < 2011] <- 2005 # Creates the two most complete veg sampling
veg <- veg %>% rename(us_species = UnderstorySp, plotid = PlotID)
unique(veg$us_species)

@@

woody_spp <- c(woody_spp, "ceanothus sp.", "cebe", "cecu", "frla", "garrya sp.", "gemo", "juca", "mafu", "mane", "mimulus sp", "pisa", "quag x quke", "qube", "quercus sp.", "quwi", "rhamnus sp.", "rhcr", "ribes sp.", "rica", "risa", "roca", "rosa sp.", "rubus sp.", "rudi", "salix sp.", "syal", "vaccinium sp.")
veg_sub <- filter(veg, veg$us_species %in% woody_spp)
veg_sub <- droplevels(veg_sub)

@@

```{r calculate overstory richness}
summary(untagged_dbh)
untagged_dbh <- as.tbl(untagged_dbh %>% select(-X) %>% rename(plotid = PlotID, species = Species))
untagged_dbh
unique(untagged_dbh$year)
untagged_dbh$year[untagged_dbh$year < 2014] <- 2005 # Create two most complete groups
untagged_dbh$plotid <- tolower(untagged_dbh$plotid)
unique(untagged_dbh$plotid)
untagged_dbh <- filter(untagged_dbh, plotid != "bush01")
untagged_dbh %>% group_by(plotid, year) %>% summarise(untagged_rich = length(unique(species)))

stems <- as.tbl(stems)
summary(stems)
unique(stems$status)
stems <- stems %>% select(-X) %>% filter(status == "Alive" | status == "Dead")
unique(stems$species)
unique(stems$plotid)
stems$species <- tolower(stems$species)
stems$species[stems$species == "quke x quag"] <- "quag x quke"
stems$plotid <- tolower(stems$plotid)
stems <- droplevels(stems)

live_tagged <- stems %>% filter(status == "Alive") %>% group_by(plotid, year, species) %>% summarise(live_count = length(species))
unique(live_tagged$species)
os_richness <- rbind(live_tagged, select(untagged_dbh, plotid, species, year, live_count))

@@

Now I want to join the richness data set and the canopy cover estimates to the `umca_plots` data frame that currently has the environmental and leaf count/infection variables.

```{r join richness and umca data}
umca_plots
richness
umca_plots <- rename(umca_plots, plotid = plot)
umca_plots$plotid <- tolower(umca_plots$plotid)

umca_plots <- left_join(umca_plots, richness, by = c("plotid", "year"))
head(canopy_cover)
canopy_cover <- select(canopy_cover, -X)
umca_plots <- left_join(umca_plots, canopy_cover, by = c("plotid", "year"))
```

@@

umca_plots <- ungroup(umca_plots)

@@

```{r join umca and qusp data}
umca_plots
plot_qusp # right now this is broken down by oak species
qusp_plots <- plot_qusp
rm(plot_qusp)
summary(qusp_plots)
qusp_plots
oak_sod <- qusp_plots %>% group_by(plot, year) %>%
      summarise(uninf_oak_ct = sum(uninfected_oak_ct), 
                inf_oak_ct = sum(infected_oak_ct), 
                tot_oak_ct = sum(c(uninfected_oak_ct, infected_oak_ct)))
oak_sod
oak_sod <- rename(oak_sod, plotid = plot)
oak_sod$plotid <- tolower(oak_sod$plotid)

oak_sod <- left_join(oak_sod, select(umca_plots, plotid, year, uninfected_bay_ct, infected_bay_ct, tot_bay, avg_lfct, tot_lfct), 
                     by = c("plotid", "year"))
summary(oak_sod) # the NA's for the bay laurel variables are legitimately zeroes b/c there are no bay in some of these plots
filter(oak_sod, is.na(tot_bay))
oak_sod$uninfected_bay_ct[is.na(oak_sod$uninfected_bay_ct)] <- 0
oak_sod$infected_bay_ct[is.na(oak_sod$infected_bay_ct)] <- 0
oak_sod$tot_bay[is.na(oak_sod$tot_bay)] <- 0

plots_env
plots_env$plotid <- tolower(plots_env$plotid)
oak_sod <- left_join(oak_sod, plots_env, by = "plotid")
head(temps_data, 15)
temps_data$plotid <- tolower(temps_data$plotid)
temps_data <- rename(temps_data, year = sample_year)
oak_sod <- left_join(oak_sod, temps_data, by = c("plotid", "year"))
oak_sod
names(oak_sod)

richness
oak_sod <- left_join(oak_sod, richness, by = c("plotid", "year"))
summary(oak_sod)
filter(oak_sod, is.na(elev15m))
oak_sod <- filter(oak_sod, plotid != "duer01", plotid != "ponti01", plotid != "sweet01")

summary(filter(oak_sod, is.na(avg_tmax)))
filter(oak_sod, is.na(avg_tmax), year == 2005) # BUSH01 plot; I'm dropping that from the data set
oak_sod <- filter(oak_sod, plotid != "bush01")
filter(oak_sod, is.na(avg_tmax), year == 2004) # PONTI01 and SWEET01 were other early abandoned plots
oak_sod <- filter(oak_sod, plotid != "ponti01", plotid != "sweet01")
summary(filter(oak_sod, is.na(avg_tmax))) # Now the NAs in the temperature variables are just for year 2003 when we had severely incomplete temperature data
summary(oak_sod)
filter(oak_sod, is.na(easting))
oak_sod <- filter(oak_sod, plotid != "duer01")

@@

oak_sod <- left_join(oak_sod, canopy_cover, by = c("plotid", "year"))
summary(rain_data)
rain_data$plotid <- tolower(rain_data$plotid)
oak_sod <- left_join(oak_sod, rain_data, by = c("plotid", "year"))
summary(filter(oak_sod, is.na(avg_tmax))) # all obs from 2003

@@

# Path Analysis Presented at 2015 IALE World Congress

#### Path Analysis Plot-Level Oak Infection 2005
Here I am using the `oak_sod` data set which will be the oak infection (binomial) as the terminal response variable.

@@

#### Path Analysis Oak Infection - 2014
Now I am going to fit the same path model to data from 2011.

@@

```{r oak path 2014 v2}
oak_path2 <- DAG(
      tmi ~ elevation,
      os_rich ~ tmi,
      hrs_14_20 ~ elevation + d2c + avg_psi + tmi,
      tot_lfct ~ hrs_14_20 + os_rich + d2c + elevation,
      inf_oak_ct ~ tot_lfct + avg_psi + os_rich + hrs_14_20 + d2c + elevation
      )
isAcyclic(oak_path2)
bu_oak_path2 <- basiSet(oak_path2)
bu_oak_path2

@@

#### Importance Value
I also want to calculate some variables related to DBH, and in particular the "importance value" for each species. Importance value is calculated in the literature as:

`IV = 0.5 * (relative density / relative dominance)` where 
 - `Relative Density = # stems of species / total # of stems` and
 - `Relative Dominance = basal area of species / total basal area`

@@

species_dbh <- rbind(tagged_dbh, untagged_dbh)
species_dbh
unique(species_dbh$species)
species_dbh %>% group_by(plotid, species, year)

stems_dbh <- species_dbh %>% group_by(plotid, year) %>% mutate(all_live_count = sum(live_count), all_live_dbh = sum(live_tot_dbh, na.rm = T), all_dead_count =  sum(dead_count, na.rm = T), all_dead_dbh = sum(dead_tot_dbh, na.rm = T))
stems_dbh
summary(stems_dbh)
filter(stems_dbh, all_live_dbh > 800)

@@

```{r pare down stems dataframe}
stems %>% select(year, tag, species, status, notes) %>% filter(species == "umca") 
summary(stems) # Limited to "Alive/Dead"" status, "Missing" statuses excluded
unique(stems$species)
unique(stems$plotid)
summary(bay_laurel) # Still includes stems with "Missing" status
unique(bay_laurel$plot)

names(stems)
stemssub <- select(stems, plotid, year, cluster, tag, species, status, slc, canker, dbh, lide_lf_symp, sod_dead, location, dbh1, dbh2, delta_dbh, dbh2_1_ratio, inst_grwth_rate, doy, date)
summary(stemssub)
str(stemssub)

summary(oak_sod) # plots with oak stems
summary(umca_plots) # plots with bay laurel stems
summary(rain_data)
summary(temps_data)
summary(plots_env)
summary(veg_sub)
summary(richness)
summary(stems_dbh)

save.image("~/GitHub/superspreaders/data_201508.RData")
```

```{r}
load("~/GitHub/superspreaders/data_201508.RData")
stemssub$cluster <- as.factor(stemssub$cluster)
stemssub$tag <- as.factor(stemssub$tag)
stemssub$canker <- as.factor(stemssub$canker)
stemssub$sod_dead <- as.factor(stemssub$sod_dead)
stemssub$date <- as.Date(stemssub$date)
str(stemssub)
```
